The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-02-28 paper/forestecology.Rmd
simon's edits to the Example section!

We demonstrate the `forestecology` package's functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot, located at the Smithsonian's National Zoo and Conservation Biology Institute in Front Royal, VA, USA @bourg_initial_2013. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US: the Blue Ridge, Ridge and Valley, and Piedmont provinces and is adjacent to the northern end of Shenandoah National Park. The forest type is typical mature secondary eastern mixed deciduous forest, with a canopy dominated by tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus* spp.), and hickories (*Carya* spp.), and an understory composed mainly of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*) @bourg_initial_2013.

@@

These two data frames are then used as arguments to the `compute_growth()` function, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`), since we are only interested in the diameter growth of surviving, and not resprouted, stems.

@@

The output `growth_scbi` is a single data frame of class `sf` that includes variables `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, and `geometry`, the `sf` package's encoding of geolocations of type `<POINT>`. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`. The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all stems is saved in a data frame of class `sf` and at a minimum the data contains the variables above. TODO: enumerate what these variables are and say if they need exact names

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, an assumption that the observations are independent underlies any cross-validation algorithm. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees. @roberts_cross-validation_2017 @pohjankukka_estimating_2017.

@@

These plots give the posterior distributions of parameters from Equation \ref{eq:model}. For many pacakge users they will be of interest because they give insight into the species-specific competitive interactions. Setting `type = "intercepts"` gives posterior distributions for $\beta_{0,j}$ and `type = "dbh_slopes"` for $\beta_{dbh,j}$. These give species specific growth independent of competition. The values of more interest are plotted with `type = 'competition'` which gives the posterior distribution for $\lambda_{j,k}$ species-specific competition coefficients (i.e., the $\lambda$-matrix). Negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip trees (`litu`) have a strong negative effect on the growth of conspecifics but relatively little effect on neighbors of the other two species.

@@

In order to evaluate the effect of competitor species identity, we use the four steps of our analysis sequence answer along with a permutation test: Under a null hypothesis where competitor species identity does not matter, we can permute/shuffle this variable within each focal tree, compute the RMSE (the test statistic of interest), repeat this process several times to construct a null distribution of the RMSE, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test randomly resamples the `comp_sp` variable with replacement, leaving all other variables intact. The resampling with replacement is nested within each focal tree in order to preserve the neighborhood structure of our competition model. To run the permutation test, we use the same`comp_bayes_lm()` function as in Section \ref{model-fit-predict}, but with a `run_shuffle = TRUE` argument.

>>>>

We demonstrate the `forestecology` package's functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot, located at the Smithsonian's National Zoo and Conservation Biology Institute in Front Royal, VA, USA @bourg_initial_2013. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park. The forest type is typical mature secondary eastern mixed deciduous forest, with a canopy dominated by tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus* spp.), and hickories (*Carya* spp.), and an understory composed mainly of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*) @bourg_initial_2013.

@@

These two data frames are then supplied as arguments to the `compute_growth()` function, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`), since we are only interested in the diameter growth of surviving, and not resprouted, stems.

@@

The output `growth_scbi` is a single data frame of class `sf` that includes variables `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, and `geometry`, the `sf` package's encoding of geolocations of type `<POINT>`. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`. The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end, all that matters is that the growth of all stems is saved in a data frame of class `sf` and, at a minimum, the data contains the variables above. TODO: enumerate what these variables are and say if they need exact names

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, many of these algorithms assume that the observations are independent of each other. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees. @roberts_cross-validation_2017 @pohjankukka_estimating_2017.

@@

These plots give the posterior distributions of parameters from Equation \ref{eq:model}. For many package users they will be of interest because they give insight into the species-specific competitive interactions. Setting `type = "intercepts"` gives posterior distributions for $\beta_{0,j}$ and `type = "dbh_slopes"` for $\beta_{dbh,j}$. These give species specific growth independent of competition. The values of more interest are plotted with `type = 'competition'` which gives the posterior distribution for $\lambda_{j,k}$ species-specific competition coefficients (i.e., the $\lambda$-matrix). Negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip trees (`litu`) have a strong negative effect on the growth of conspecifics but relatively little effect on neighbors of the other two species.

@@

In order to evaluate the effect of competitor species identity, we use the four steps of our analysis sequence answer along with a permutation test: Under a null hypothesis where competitor species identity does not matter, we can permute/shuffle this variable within each focal tree, compute the RMSE (the test statistic of interest), repeat this process several times to construct a null distribution of the RMSE, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test randomly resamples the `comp_sp` variable with replacement, leaving all other variables intact. The resampling with replacement is nested within each focal tree in order to preserve the neighborhood structure of our competition model. To run the permutation test, we use the same `comp_bayes_lm()` function as in Section \ref{model-fit-predict}, but with a `run_shuffle = TRUE` argument.
